---
title: "Noise_Testing"
format: html
---

## Noise Implementation

```{r}
set.seed(123)
r <- 4; p <- 4
simulate_group <- function(n, mean_matrix, row_sd = 0.5, col_sd = 0.5) {
	row_cov <- diag(row_sd, nrow(mean_matrix))
	col_cov <- diag(col_sd, ncol(mean_matrix))
	lapply(seq_len(n), function(i) mean_matrix + row_cov %*% matrix(rnorm(r * p), r, p) %*% col_cov)
}

M1 <- matrix(1, r, p)
M2 <- matrix(-1, r, p)
x_list <- c(simulate_group(40, M1), simulate_group(20, M2))

contam <- lapply(seq_len(15), function(i) matrix(runif(r * p, -5, 5), r, p))
x_list <- c(x_list, contam)
```

```{r}
library(ggplot2)

X <- t(sapply(x_list, c))

group <- factor(c(rep("M1", 40),
                  rep("M2", 20),
                  rep("Contam", 15)))

pca <- prcomp(X, scale. = TRUE)

df <- data.frame(
  PC1 = pca$x[,1],
  PC2 = pca$x[,2],
  group = group
)

ggplot(df, aes(PC1, PC2, color = group)) +
  geom_point(size = 3, alpha = 0.8) +
  theme_minimal()
```


### HC-noise

```{r}
library(Ampharos)
fit_hc <- Ampharos::matrix_variate_noise_fit(x_list, g = 2, noise_type = "hc",
                                             estimate_k = TRUE,
                                             verbose = TRUE)

#fit_hc <- Ampharos::matrix_noise_auto_k_robust(x_list, g = 2, verbose = TRUE)
# fit_hc <-  Ampharos::matrix_variate_noise_fit(x_list, g = 2, noise_type = "hc", noise_k = 1e-09, verbose = FALSE)

cat('HC fit mixing proportions:\n')
print(fit_hc$pi)
cat('\nHC cluster counts (0 indicates noise):\n')
print(table(fit_hc$cluster))
print(fit_hc$noise)
```

```{r}
k_grid <- 10^seq(-15, -1, length.out = 30)

for(i in seq_along(k_grid)){
  fit_hc <- Ampharos::matrix_variate_noise_fit(x_list, g = 3, noise_type = "hc", 
                                             noise_k = k_grid[i], 
                                             verbose = FALSE)
Ampharos::matrix_noise_ecdf_vs_cdf_plot(
  fit_hc_viroli,
  x_list,
  main = "HC Noise Fit: ECDF vs Chi-square CDF"
)
}
```


### BR-noise

```{r}
fit_br <- matrix_variate_noise_fit(x_list, g = 2, noise_type = "br", 
                                   nstart = 100, max_iter = 50, verbose = FALSE)

cat('BR fit mixing proportions:\n')
print(fit_br$pi)
cat('\nBR cluster counts (0 indicates noise):\n')
print(table(fit_br$cluster))
print(fit_br$noise)
```

HC-noise recovered the two clusters and the noise proportion accurately.

BR fit largely collapsed the two true clusters into one component (component_2), leaving component_1 with very few items. Noise proportion is near the true value (~0.13), but cluster recovery failed for the two Gaussians.

Box diagnostics: per-entry limits $\approx$ +-13–15 and log-volume $\approx$ 40 nears uniform density = exp(-40) (extremely small). That tiny uniform density makes the BR background numerically negligible compared with Gaussian peaks, which can change how the EM balances components and makes results sensitive to initialization.
Likely causes for the BR behaviour

Bounding-box volume is huge (because contamination inflates min/max), so the BR uniform density is extremely small and competes poorly with Gaussians; EM may then prefer to explain almost-all data with one Gaussian and use the other Gaussian only for a handful of points.
Initialization sensitivity (kmeans nstart small) can let components merge.
BR using the simple axis-aligned box is fragile when contamination sets wide margins.

## Simulation in the Style of Viroli

A total of 100 datasets are created with n = 300 $3 \times 5$ matrices in three groups with proportions $\pi = (0.3, 0.4, 0.3)$ and means of the form

$$
M_g = \begin{pmatrix}
    a & 0 & 0 & 0 & 0\\
    b & 0 & 0 & 0 & 0\\
    0 & 0 & 0 & 0 & 0
\end{pmatrix}
$$

with M_1 : $(a,b)$ = $(0.5, 0.5)$, M_2 : $(a,b)$ = $(0,0)$ and M_3 : $(a,b)$ = $(-0.5, 0.5)$.

```{r}
library(clusterGeneration)

set.seed(42)
r_viroli <- 3; p_viroli <- 5
n_viroli <- 300

# Viroli-style mean matrices (3x5 with first column containing a,b values)
M1_viroli <- matrix(c(0.5, 0.5, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0), nrow = r_viroli, ncol = p_viroli, byrow = FALSE)
M2_viroli <- matrix(0, r_viroli, p_viroli)
M3_viroli <- matrix(c(-0.5, 0.5, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0), nrow = r_viroli, ncol = p_viroli, byrow = FALSE)

# Generate random U and V matrices
U1_viroli <- rcorrmatrix(r_viroli)
V1_viroli <- rcorrmatrix(p_viroli)
U2_viroli <- rcorrmatrix(r_viroli)
V2_viroli <- rcorrmatrix(p_viroli)
U3_viroli <- rcorrmatrix(r_viroli)
V3_viroli <- rcorrmatrix(p_viroli)

# Simulate groups with Viroli proportions (0.3, 0.4, 0.3)
n1_viroli <- round(0.3 * n_viroli)
n2_viroli <- round(0.4 * n_viroli)
n3_viroli <- n_viroli - n1_viroli - n2_viroli

simulate_matrix_group <- function(n, mean_mat, row_cov, col_cov) {
  lapply(seq_len(n), function(i) {
    mean_mat + row_cov %*% matrix(rnorm(r_viroli * p_viroli), r_viroli, p_viroli) %*% col_cov
  })
}

x_list_viroli <- c(
  simulate_matrix_group(n1_viroli, M1_viroli, U1_viroli, V1_viroli),
  simulate_matrix_group(n2_viroli, M2_viroli, U2_viroli, V2_viroli),
  simulate_matrix_group(n3_viroli, M3_viroli, U3_viroli, V3_viroli)
)

# Add permuted outliers (15 matrices with permuted entries)
outlier_idx_viroli <- sample(length(x_list_viroli), 15)
for (idx in outlier_idx_viroli) {
  x_list_viroli[[idx]] <- matrix(sample(x_list_viroli[[idx]]), r_viroli, p_viroli)
}

cat('Viroli simulation: n =', length(x_list_viroli), ', groups =', n1_viroli, n2_viroli, n3_viroli, '\n')
```

### Viroli: HC-noise with Automatic K

```{r}
fit_hc_viroli <- Ampharos::matrix_variate_noise_fit(x_list_viroli, g = 3, noise_type = "hc",
                                                     noise_k = TRUE,
                                                     verbose = TRUE)


# fit_hc_viroli <- Ampharos::matrix_variate_noise_fit(x_list_viroli, g = 3, noise_type = "hc",
#                                                      noise_k = 1.709065e-09,
#                                                      verbose = TRUE)
# 

cat('HC fit mixing proportions:\n')
print(fit_hc_viroli$pi)
cat('\nHC cluster counts (0 indicates noise):\n')
print(table(fit_hc_viroli$cluster))
cat('\nSelected k:', fit_hc_viroli$k_selection$selected_k, '\n')
```

### Viroli: PCA Visualization

```{r}
# Extract matrix data for visualization
matrices_array <- array(NA, dim = c(r_viroli, p_viroli, length(x_list_viroli)))
for(i in 1:length(x_list_viroli)) {
  matrices_array[,,i] <- x_list_viroli[[i]]
}

# Reshape each matrix into a vector for PCA
matrix_vectors <- t(apply(matrices_array, 3, function(x) as.vector(t(x))))

# Perform PCA
pca_result <- prcomp(matrix_vectors, center = TRUE, scale. = FALSE)
plot_df <- data.frame(
  PC1 = pca_result$x[,1],
  PC2 = pca_result$x[,2],
  Cluster = factor(fit_hc_viroli$cluster, 
                   levels = c(0,1,2,3),
                   labels = c("Noise", "Group 1", "Group 2", "Group 3")),
  True_Group = factor(rep(c(1,2,3), times = c(n1_viroli, n2_viroli, n3_viroli)))
)

outlier_flag <- rep(FALSE, length(x_list_viroli))
outlier_flag[outlier_idx_viroli] <- TRUE

# Plot 1: PCA colored by assigned cluster
library(ggplot2)
p1 <- ggplot(plot_df, aes(x = PC1, y = PC2, color = Cluster)) +
  geom_point(alpha = 0.8, size = 2) +
  theme_minimal() +
  labs(title = "PCA of Matrix Data (Colored by HC Cluster Assignment)",
       subtitle = paste0("n=", n_viroli, " | Noise=", sum(fit_hc_viroli$cluster==0),
                         " | Selected k=", format(fit_hc_viroli$k_selection$selected_k, scientific=TRUE))) +
  theme(legend.position = "bottom") +
  theme_classic() + 
  scale_color_manual(values = c("Noise" = "black", 
                                "Group 1" = "red", 
                                "Group 2" = "green", 
                                "Group 3" = "blue"))

print(p1)
```

## Simulation in the Style of Tomarchio

A total of 100 datasets are created with n = 200 $2 \times 4$ matrices in two groups of equal weight.

$$
\begin{aligned}
M_1 = \begin{pmatrix}
      -2.60 & -1.10 & -0.50 & -0.20\\
      1.30 & 0.60 & 0.30 & 0.10
    \end{pmatrix}
\qquad
&M_2 = \begin{pmatrix}
      1.50 & 1.70 & 1.90 & 2.20\\
      -3.70 & -2.70 & -2.00 & -1.50
    \end{pmatrix} \\

U_1 = \begin{pmatrix}
      2 & 0 \\
      0 & 1
    \end{pmatrix}
\qquad
&U_2 = \begin{pmatrix}
      1.70 & 0.5 \\
      0.5 & 1.30
    \end{pmatrix}
\end{aligned} \\

V_1 = V_2 = \begin{pmatrix}
      1.00 & 0.50 & 0.25 & 0.13 \\
      0.50 & 1.00 & 0.50 & 0.25 \\
      0.25 & 0.50 & 1.00 & 0.50 \\
      0.13 & 0.25 & 0.50 & 1.00
    \end{pmatrix}
$$

Ten matrices are randomly selected, and then one column from each selected matrix is randomly replaced with data generated from Unif(-15, 15).

```{r}
set.seed(123)
r_tomarchio <- 2; p_tomarchio <- 4
n_tomarchio <- 200

# Tomarchio mean matrices
M1_tomarchio <- matrix(c(-2.60, -1.10, -0.50, -0.20, 1.30, 0.60, 0.30, 0.10), nrow = r_tomarchio, ncol = p_tomarchio, byrow = FALSE)
M2_tomarchio <- matrix(c(1.50, 1.70, 1.90, 2.20, -3.70, -2.70, -2.00, -1.50), nrow = r_tomarchio, ncol = p_tomarchio, byrow = FALSE)

# Tomarchio covariance matrices
U1_tomarchio <- matrix(c(2, 0, 0, 1), nrow = r_tomarchio, ncol = r_tomarchio)
U2_tomarchio <- matrix(c(1.70, 0.5, 0.5, 1.30), nrow = r_tomarchio, ncol = r_tomarchio)

V_tomarchio <- matrix(c(1.00, 0.50, 0.25, 0.13,
                        0.50, 1.00, 0.50, 0.25,
                        0.25, 0.50, 1.00, 0.50,
                        0.13, 0.25, 0.50, 1.00), nrow = p_tomarchio, ncol = p_tomarchio)

# Simulate groups with equal proportions
n1_tomarchio <- round(0.5 * n_tomarchio)
n2_tomarchio <- n_tomarchio - n1_tomarchio

simulate_tomarchio_group <- function(n, mean_mat, row_cov, col_cov) {
  lapply(seq_len(n), function(i) {
    mean_mat + row_cov %*% matrix(rnorm(r_tomarchio * p_tomarchio), r_tomarchio, p_tomarchio) %*% col_cov
  })
}

x_list_tomarchio <- c(
  simulate_tomarchio_group(n1_tomarchio, M1_tomarchio, U1_tomarchio, V_tomarchio),
  simulate_tomarchio_group(n2_tomarchio, M2_tomarchio, U2_tomarchio, V_tomarchio)
)

# Add column contamination outliers (10 matrices with one column replaced by Unif(-15, 15))
contam_idx_tomarchio <- sample(length(x_list_tomarchio), 10)
for (idx in contam_idx_tomarchio) {
  col_to_replace <- sample(p_tomarchio, 1)
  x_list_tomarchio[[idx]][, col_to_replace] <- runif(r_tomarchio, -15, 15)
}

cat('Tomarchio simulation: n =', length(x_list_tomarchio), ', groups =', n1_tomarchio, n2_tomarchio, '\n')
```

### Tomarchio: HC-noise with Automatic K

```{r}
fit_hc_tomarchio <- Ampharos::matrix_variate_noise_fit(x_list_tomarchio, g = 2, noise_type = "hc",
                                                      estimate_k = TRUE,
                                                      verbose = TRUE)

cat('HC fit mixing proportions:\n')
print(fit_hc_tomarchio$pi)
cat('\nHC cluster counts (0 indicates noise):\n')
print(table(fit_hc_tomarchio$cluster))
cat('\nSelected k:', fit_hc_tomarchio$k_selection$selected_k, '\n')
```


