---
title: "Noise_Testing"
format: html
---

## Noise Implementation

```{r}
set.seed(123)
r <- 4; p <- 4
simulate_group <- function(n, mean_matrix, row_sd = 0.5, col_sd = 0.5) {
	row_cov <- diag(row_sd, nrow(mean_matrix))
	col_cov <- diag(col_sd, ncol(mean_matrix))
	lapply(seq_len(n), function(i) mean_matrix + row_cov %*% matrix(rnorm(r * p), r, p) %*% col_cov)
}

M1 <- matrix(1, r, p)
M2 <- matrix(-1, r, p)
x_list <- c(simulate_group(40, M1), simulate_group(20, M2))

contam <- lapply(seq_len(15), function(i) matrix(runif(r * p, -6, 8), r, p))
x_list <- c(x_list, contam)
```

### HC-noise

```{r}
library(Ampharos)
k_grid <- 10^seq(-200, -1, length.out = 30)
fit_hc <- Ampharos::matrix_variate_noise_fit(x_list, g = 2, noise_type = "hc",
                                             estimate_k = TRUE,
                                             k_grid = k_grid,
                                             verbose = TRUE)

#fit_hc <- Ampharos::matrix_noise_auto_k_robust(x_list, g = 2, verbose = TRUE)

# fit_hc <-  Ampharos::matrix_variate_noise_fit(x_list, g = 2, noise_type = "hc", noise_k = 1e-09, verbose = FALSE)

cat('HC fit mixing proportions:\n')
print(fit_hc$pi)
cat('\nHC cluster counts (0 indicates noise):\n')
print(table(fit_hc$cluster))
print(fit_hc$noise)
```

```{r}
k_grid <- 10^seq(-200, -1, length.out = 30)
for(i in seq_along(k_grid)){
  fit_hc <- Ampharos::matrix_variate_noise_fit(x_list, g = 2, noise_type = "hc", 
                                             noise_k = k_grid[i], 
                                             verbose = FALSE)
Ampharos::matrix_noise_ecdf_vs_cdf_plot(
  fit_hc,
  x_list,
  main = "HC Noise Fit: ECDF vs Chi-square CDF"
)
}
```


### BR-noise

```{r}
fit_br <- matrix_variate_noise_fit(x_list, g = 2, noise_type = "br", 
                                   nstart = 100, max_iter = 50, verbose = FALSE)

cat('BR fit mixing proportions:\n')
print(fit_br$pi)
cat('\nBR cluster counts (0 indicates noise):\n')
print(table(fit_br$cluster))
print(fit_br$noise)
```

HC-noise recovered the two clusters and the noise proportion accurately.

BR fit largely collapsed the two true clusters into one component (component_2), leaving component_1 with very few items. Noise proportion is near the true value (~0.13), but cluster recovery failed for the two Gaussians.

Box diagnostics: per-entry limits $\approx$ +-13–15 and log-volume $\approx$ 40 nears uniform density = exp(-40) (extremely small). That tiny uniform density makes the BR background numerically negligible compared with Gaussian peaks, which can change how the EM balances components and makes results sensitive to initialization.
Likely causes for the BR behaviour

Bounding-box volume is huge (because contamination inflates min/max), so the BR uniform density is extremely small and competes poorly with Gaussians; EM may then prefer to explain almost-all data with one Gaussian and use the other Gaussian only for a handful of points.
Initialization sensitivity (kmeans nstart small) can let components merge.
BR using the simple axis-aligned box is fragile when contamination sets wide margins.

```{r}
# Vectorize matrices
X <- do.call(
  rbind,
  lapply(x_list, as.vector)
)

# Labels for plotting
group <- c(
  rep("Cluster 1", 40),
  rep("Cluster 2", 20),
  rep("Noise", 5)
)

# PCA
pca <- prcomp(X, center = TRUE, scale. = TRUE)

plot(
  pca$x[, 1],
  pca$x[, 2],
  col = c(
    "dodgerblue",
    "firebrick",
    "black"
  )[factor(group)],
  pch = c(
    16,
    17,
    8
  )[factor(group)],
  xlab = "PC1",
  ylab = "PC2",
  main = "PCA of Matrix-Variate Data"
)

legend(
  "topright",
  legend = c("Cluster 1", "Cluster 2", "Noise"),
  col = c("dodgerblue", "firebrick", "black"),
  pch = c(16, 17, 8),
  bty = "n"
)
```

